Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-14 appendix.tex
updating

\section{Applicability of the vertically global shearing box}\label{global_corr}
In \S\ref{analytical} we effectively employed the vertically global shearing box
\citep[VGSB,][]{mcnally14} to make analytic progress.   
The main difference between the linear problem in the VGSB 
and the radially-local approximation
(Eq. \ref{lin_mass}---\ref{lin_energy}) adopted in our numerical study
is the absence of terms explicitly associated with the radial disk structure in
the former framework. These terms appear in the energy, mass and radial momentum
equations and are labeled by $\hat{g}_c$ in
Eq. \ref{lin_mass}---\ref{lin_energy}. Nominally $\hat{g}_c=1$.

@@

\subsection{Spurious growth of adiabatic perturbations in the VGSB}\label{analytic_adia} 
An important caveat of the VGSB is that it cannot be used in the
adiabatic limit. We explain this by considering vertically isothermal
disks and setting  $\beta\to\infty$ in 
Eq. \ref{vertiso_gov_nondim} (which itself is obtained by setting
$\hat{g}_c=0$ in Eq. \ref{lin_mass}---\ref{lin_energy}) to give 
\begin{align}
  0 =\dd v_z^{\prime\prime} + \left(1 + \ii \epsilon q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime
  +\left\{\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) 
    -\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii\epsilon  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}
We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  \hat{\sigma}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat 
  =&  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
&+
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii\epsilon
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat
+ \ii\epsilon q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
% When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
% $\hat{\sigma}^2$ is real, and  $\hat{\sigma}^2>0$ if $\gamma>1$. As
% expected, a sub-adiabatically stratified disk is stable in the absence
% of vertical shear.

In the presence of vertical shear $q\neq0$, Eq. \ref{adia_integral}
suggests $\hat{\sigma}$ is complex for real $\khat$, which implies the
possibility of instability, regardless of the value of
$|\gamma-1|$. This is contradicts the second Solberg-Hoiland criterion 
(Eq. \ref{solberg2}), which imply instability requires 
$|\gamma-1|\ll1$ for thin, vertically isothermal disks evolved adiabatically. 
%also need real freq^2 in adia limit 
This is because the VGSB retains the vertical shear
($\p_z\Omega$) from the global disk, but ignores the global radial
disk structure responsible for it. Nevertheless, we demonstrate below
that this inconsistency is unimportant for the VSI because it occurs
for $\beta\ll1$.

@@

All three methods give similar behavior, and growth rates are in close
agreement for $\beta\lesssim 1$. Differences arise for 
$\beta\gtrsim1$, and as $\beta\to\infty$ the VGSB framework gives a
(spurious) growth rate as expected from the discussion
above. Inclusion of the global radial gradient terms results in the
correct behavior  ($\nu\to0$ as $\beta\to\infty$). Despite this
caveat, Fig. \ref{gcorr_compare} shows that provided we consider
$\beta\ll1$, then the VGSB framework is an adequate approximation to
study the VSI.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0cm]{figures/gcorr_compare} 
  \caption{Growth rate of the fundamental VSI mode as a function of
    the thermal relaxation time $\beta$. The `Analytical (VGSB)' curve
    is calculated from the dispersion relation Eq. \ref{relax_disp};
    the `Numerical (VGSB)' curve is obtained by numerically solving
    the linear problem, Eq. \ref{lin_mass}---\ref{lin_energy}, with $\hat{g}_c=0$
    (which defines the linear VGSB equations); and the `Numerical
    (Full)' curve is obtained by numerically solving 
    Eq. \ref{lin_mass}---\ref{lin_energy}  with $\hat{g}_c=1$, which
    accounts for the radial disk structure.  
    The disk parameters are $\Gamma=1.011$,
    $\gamma=1.4$ and $(p,q,\epsilon)=(-1.5,-1,0.05)$. The perturbation
    wavenumber is $\khat=30$. 
    \label{gcorr_compare}}  
\end{figure}

@@

In the global disk with $\Gamma=1$ we have
\begin{align}\label{dkappa2}
  \frac{\p\kappa^2}{\p z} = 4 \frac{\p\Omega^2}{\p z} + r\frac{\p}{\p
    r}\frac{\p\Omega^2}{\p z} = -
  \frac{\p\ln\rho}{\p z}\frac{qc_s^2}{r^2} \left(2 + q +
    \frac{z^2/r^2-2}{z^2/r^2+1}\right). 
\end{align}
For the local problem we may then write
\begin{align}
  \frac{dD}{dz}  = - \frac{d\ln\rho}{dz}\frac{qc_s^2}{r^2}F(z;q),
\end{align}
where the function $F$ corresponds to the bracket in
Eq. \ref{dkappa2}. This function varies from $F=q$ at $z=0$ to $F\to
3+q$ as $|z|\to\infty$. Since we consider $q=O(1)$, $|F|$ is of
order unity.

Now, eliminating $W$ and $Q$ from Eq. \ref{ode_w}---\ref{ode_Q} and
Eq. \ref{lin_vz} for vertically isothermal disks, retaining the
vertical dependence of $D$ and including thermal relaxation, we
obtain 
\begin{align}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left[1 + \frac{\ii k_x c_s^2
      q}{Dr} - \frac{k_x^2c_s^2}{\left(k_x^2c_s^2 + \chi
        D\right)}\frac{qc_s^2F}{Dr^2}\right]\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left\{\sigma^2\left(\frac{k_x^2}{D} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{Dr}\right)\frac{d^2\ln\rho}{dz^2} -
    \frac{c_s^2}{D}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left[\left(1-\chi\right) +
     \frac{\chi}{\left(k_x^2c_s^2 + \chi D\right)}\frac{qc_s^2 F}{r^2}\right] 
   \right\}\delta v_z,
\end{align}
where we recall $\chi = \left(1-\ii\sigma t_c\right)/\left(1-\ii\sigma
t_c \gamma\right)$. Making the low-frequency approximation with the
replacement $D\to \Omega_k^2$ gives, in terms of dimensionless
variables,
\begin{align}
   &\delta v_z ^{\prime\prime} + \left[1 + \ii\epsilon q\hat{k} -
    \frac{ \hat{k}^2}{
      \left(\hat{k}^2+\chi\right)}q \epsilon^2F\right]\ln\rho^{\prime}\delta v_z^\prime +
  \left\{\left(\chi + \ii \epsilon q
      \hat{k}\right)\ln\rho^{\prime\prime} - \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii\epsilon
      q\hat{k}\right)\left[1 - \chi +
      \frac{\chi}{\left(\hat{k}^2+\chi\right)}q\epsilon^2F\right]\right\}\delta v_z \notag\\&=
  -\hat{\sigma}^2\left(\hat{k}^2+\chi\right)\delta v_z\label{adia_diso3}.
\end{align}   
Eq. \ref{adia_diso3} differs from Eq. \ref{vertiso_gov_nondim} by terms
proportional to $\epsilon^2$. For a thin disk, $\epsilon\ll1$, so
neglecting these terms has no qualitative effect on the discussion in 
\S\ref{analytic_relax}.  % This issue does not
% arise for isothermal perturbations discussed in \S\ref{iso_discuss}
% since in that case we work with a governing equation for $W$ instead.

@@

\subsection{Limiting form with real frequencies and $\khat^2\gg 1$}\label{disp_neut_limit}
Assuming neutral modes with real $\hat{\sigma}=\hat{\omega}$ and $\khat^2\gg 1$, the real and imaginary parts of
Eq. \ref{relax_disp} are approximately 
\begin{align}
  0 =&M(M+1)(1-\epsilon^2 q^2 \khat^2) + 4\beta\epsilon q M(M+1)\left(\hat{\omega}\khat\right) 
 +\left[1 +
    \gamma\beta^2\left(1-\gamma\right)(1+2M)^2\right]\left(\hat{\omega}\khat\right)^2
 + 2\epsilon q \gamma\beta \left(\hat{\omega}\khat\right)^3 -  \left(\hat{\omega}\khat\right)^4\notag \\
  &+ \beta^2\gamma^2\hat{\omega}^2
  \left(\hat{\omega}\khat\right)^4,\label{relax_cond1} \\
   0=& 2\epsilon q M (M+1)\khat +
   \beta(1-\gamma)(1+2M)^2\khat\left(\hat{\omega}\khat\right) 
   + \epsilon q \khat \left(\hat{\omega}\khat\right)^2 -
   2\beta\hat{\omega}\left(\hat{\omega}\khat\right)^2
   - \epsilon q
   \gamma^2\beta^2\hat{\omega}\left(\hat{\omega}\khat\right)^3 
   +
   2\beta\gamma\hat{\omega}\left(\hat{\omega}\khat\right)^4 \label{relax_cond2}. 
\end{align}
These equations are used to derive a critical thermal timescale in
\S\ref{iso_vsi_beta_crit}. For fixed $M$ of $O(1)$,  we assume
$|\hat{\omega}\khat|$ remain constant but consider $\khat\gg 1$ so
that $|\hat{\omega}|\ll 1$.  Then for a thin disk $\epsilon \ll 1$
we neglect the second, fourth and last term in Eq. \ref{relax_cond1};
and neglect the last three terms in Eq. \ref{relax_cond2}. This
procedure gives Eq. \ref{relax_cond_simp1}---\ref{relax_cond_simp2},
but is not valid if $\beta$ or $|q|$ is large.

@@

%  In this case the fourth term
% in Eq. \ref{relax_cond1} may be retained but other aforementioned terms
% are still neglected. One then obtains an expression for $\beta$ by
% eliminating $\hat{\omega}\khat$ between Eq. \ref{relax_cond1} ---
% \ref{relax_cond2}. The expression for the critical thermal timescale
% for the fundamental mode, $\beta_\mathrm{crit}$
% (Eq. \ref{iso_vsi_con}, is then obtained by taking the thin disk
% limit.

@@

\subsubsection{Maximum growth rate}\label{max_growth1}
Here we show that the growth rate is limited by the maximum vertical
shear in the domain. % Writing $\hat{\sigma} = \hat{\omega} +
% \ii\hat{\nu}$ with real $\hat{\omega}$ and $\hat{\nu}$, 
The real and imaginary parts of 
Eq. \ref{integral_relation1} are
\begin{align}
  \left(\hat{\omega}^2-\hat{\nu}^2\right)\left(1+\hat{k}^2\right)
  \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z} -
  \int_{\zhat_1}^{\zhat_2}\rho|W^\prime|^2d\hat{z}
  \notag
  =\real\left[\ii
    \epsilon\hat{k}q \int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\zhat\right], \\
   2\hat{\omega}\hat{\nu}\left(1+\hat{k}^2\right)
  \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z}
  =\imag\left[\ii
    \epsilon\hat{k}q \int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right],
\end{align}
where we recall $\hat{\omega} = \real \hat{\sigma}$ and
$\hat{\nu}=\imag\hat{\sigma}$. 
Adding the square of these equations give
\begin{align}
  \left[|\hat{\sigma}|^2\left(1+\hat{k}^2\right)
    \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z} -
    \int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}\right]^2
  +4\hat{\nu}^2\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}
  =\left|\ii
    \epsilon\hat{k}q\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|^2.
\end{align}
It is clear that
\begin{align}\label{sigma_finite_domain} 
  4\hat{\nu}^2\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime
  \right|^2d\hat{z} 
  \leq\left|
    \epsilon\hat{k}q\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|^2.
\end{align}
On the left hand side of this inequality, we apply the Cauchy-Schwarz
inequality to obtain
\begin{align}
  \left( \int_{\zhat_1}^{\zhat_2}\rho
    |W|\left|W^\prime \right|d\hat{z}\right)^2\leq
  \int_{\zhat_1}^{\zhat_2}\rho 
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}.
\end{align}
On the right hand side of Eq. \ref{sigma_finite_domain} we have
\begin{align}
  \left|\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|\leq \int_{\zhat_1}^{\zhat_2}\rho
  \left|f(\hat{z})W^*W^\prime \right|d\hat{z}
  \leq
  \mathrm{max}\left(|f|\right)\int_{\zhat_1}^{\zhat_2}\rho
  |W|\left|W^\prime \right|d\hat{z},
\end{align}
where $\mathrm{max}(|f|)$ is the maximum value of $|f|$ in
$\zhat\in[\zhat_1,\zhat_2]$. Inserting these inequalities into
Eq. \ref{sigma_finite_domain} gives
\begin{align}\label{max_growth}
  |\hat{\nu}|\leq
  \frac{\epsilon |\hat{k} q|}{2\sqrt{1+\hat{k}^2}}\mathrm{max}(|f|). 
\end{align}
Recalling that $f$ represents vertical shear (Eq. \ref{fz_shear}), it
follows that the maximum possible growth rate of unstable modes,
satisfying the above boundary conditions, is limited by the maximum
vertical shear in the domain considered.

@@

We seek power series solutions to Eq. \ref{iso_ode3}, 
\begin{align}
  W(\zhat) = \sum_{l=0}^\infty a_l\zhat^l. 
\end{align}
Then the coefficients must satisfy the recurrence relation
\begin{align}
  (l+2)(l+1)a_{l+2} +
  \left[\hat{\sigma}^2\left(1+\frac{\khat^2}{1-\hat{\sigma}^2}\right)
    - l\left(1+\frac{\ii \epsilon q
        \hat{k}}{1-\hat{\sigma}^2}\right)\right] a_l = 0. 
\end{align}
We demand the series to terminate  at $l=L$, i.e. a polynomial
solution, so that the vertical kinetic energy remain bounded as
$|\zhat|\to\infty$.  Then the eigenfrequency $\hat{\sigma}$ is given via 
\begin{align}
\hat{\sigma}^4 - \left(L+1+\khat^2\right)\hat{\sigma}^2 + L\left(1 +
  \ii\epsilon q \khat\right) = 0. \label{simple_growth}
\end{align}

\subsubsection{Stability without vertical shear}\label{stable_novshear}
In the absence of vertical shear $q=0$, Eq. \ref{simple_growth} can be
written as \begin{align}
  \hat{\sigma}^2\khat^2 =
  \left(L-\hat{\sigma}^2\right)\left(1-\hat{\sigma}^2\right). \label{iso_disp_full}
\end{align}
This is just the dispersion relation for axisymmetric isothermal waves in a
vertically isothermal disk
\citep[e.g.][]{okazaki87,takeuchi98,tanaka02,zhang06,ogilvie13,barker14}. In 
this case the solutions are Hermite polynomials, $W\propto
\He_L(\zhat)$.  The eigenfrequency $\hat{\sigma} = \hat{\omega}$ is real and the disk
is stable. The low-frequency branch of Eq. \ref{iso_disp_full} are 
inertial waves \citep{balbus03}. For 
$|\hat{\omega}|\ll 1$ and $L\geq 1$ the dispersion relation is
$\hat{\omega}^2\khat^2 = L$, or $\hat{\omega}\propto \khat^{-1}$ for
fixed $L$.  This inverse relation has been qualitatively
observed in numerical simulations of \cite{stoll14}.

>>>>

\section{Applicability of the vertically global shearing box}\label{global_corr}
In \S\ref{analytical} we effectively employed the vertically global shearing box
\citep[VGSB,][]{mcnally14} to make analytic progress.   
The difference between the linear problem in the VGSB 
and the radially-local approximation adopted in our numerical study
is the absence of terms explicitly associated with the radial disk structure in
the VGSB. These terms appear in the energy, mass and radial momentum
equations and are labeled by $\hat{g}_c$ in
Eq. \ref{lin_mass}---\ref{lin_energy}. Nominally $\hat{g}_c=1$.

@@

\subsection{Spurious growth of adiabatic perturbations in the VGSB}\label{analytic_adia} 
An important caveat of the VGSB is that it cannot be used in the
adiabatic limit. We explain this by considering vertically isothermal
disks and setting  $\beta\to\infty$ in 
Eq. \ref{vertiso_gov_nondim} (which itself is obtained by setting
$\hat{g}_c=0$ in Eq. \ref{lin_mass}---\ref{lin_energy}) to give 
\begin{align}
  0 =\dd v_z^{\prime\prime} + \left(1 + \ii \epsilon q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime
  +\left\{\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) 
    -\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii\epsilon  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}
We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  \hat{\sigma}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat 
  =&  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
&+
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii\epsilon
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat
+ \ii\epsilon q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
% When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
% $\hat{\sigma}^2$ is real, and  $\hat{\sigma}^2>0$ if $\gamma>1$. As
% expected, a sub-adiabatically stratified disk is stable in the absence
% of vertical shear. 
In the presence of vertical shear $q\neq0$, Eq. \ref{adia_integral}
suggests $\hat{\sigma}^2$ is complex for real $\khat$, which implies the
possibility of instability, regardless of the value of
$|\gamma-1|$. This is contradicts the second Solberg-Hoiland criterion 
(Eq. \ref{solberg2}), which imply instability for adiabatic
perturbations in a thin, vertically isothermal disk would require 
$|\gamma-1|\ll1$. 
%also need real freq^2 in adia limit 
This is because the VGSB retains the vertical shear
($\p_z\Omega$) from the global disk, but ignores the global radial
disk structure responsible for it. Nevertheless, we demonstrate below
that this inconsistency is unimportant for the VSI because it occurs
for $\beta\ll1$.

@@

All three methods give similar behavior, and growth rates are in close
agreement for $\beta\lesssim 1$. Differences arise for 
$\beta\gtrsim1$, and as $\beta\to\infty$ the VGSB framework gives a
(spurious) growth rate as expected from the discussion
above. Inclusion of the global radial gradient terms results in the
correct behavior  ($\nu\to0$ as $\beta\to\infty$). Despite this
caveat, Fig. \ref{gcorr_compare} shows that provided we consider
$\beta\ll1$, then the VGSB framework is adequate for studying the
VSI.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0cm]{figures/gcorr_compare} 
  \caption{Growth rate of the fundamental VSI mode as a function of
    the thermal relaxation time $\beta$. The `Analytical (VGSB)' curve
    is calculated from the dispersion relation Eq. \ref{relax_disp};
    the `Numerical (VGSB)' curve is obtained by numerically solving
    the linear problem, Eq. \ref{lin_mass}---\ref{lin_energy}, with $\hat{g}_c=0$
    (which defines the linear VGSB equations); and the `Numerical
    (Full)' curve is obtained by numerically solving 
    Eq. \ref{lin_mass}---\ref{lin_energy}  with $\hat{g}_c=1$, which
    accounts for the radial disk structure.  
    The disk parameters are $(\gamma, \Gamma)=(1.4, 1.011)$ and
    $(p,q,\epsilon)=(-1.5,-1,0.05)$. The perturbation
    wavenumber is $\khat=30$. 
    \label{gcorr_compare}}  
\end{figure}

@@

In the global disk with $\Gamma=1$ we have
\begin{align}\label{dkappa2}
  \frac{\p\kappa^2}{\p z} = 4 \frac{\p\Omega^2}{\p z} + r\frac{\p}{\p
    r}\frac{\p\Omega^2}{\p z} = -
  \frac{\p\ln\rho}{\p z}\frac{qc_s^2}{r^2} \left(2 + q +
    \frac{z^2/r^2-2}{z^2/r^2+1}\right). 
\end{align}
For the local problem we may then write
\begin{align}
  \frac{dD}{dz}  = - \frac{d\ln\rho}{dz}\frac{qc_s^2}{r^2}F(z;q),
\end{align}
where the function $F$ corresponds to the bracket in
Eq. \ref{dkappa2}. This function varies from $F=q$ at $z=0$ to $F\to
3+q$ as $|z|\to\infty$. Since we consider $|q|=O(1)$, this impies
$|F|=O(1)$ as well.

Now, eliminating $W$ and $Q$ from Eq. \ref{ode_w}---\ref{ode_Q} and
Eq. \ref{lin_vz} for vertically isothermal disks, retaining the
vertical dependence of $D$ and including thermal relaxation, we
obtain 
\begin{align}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left[1 + \frac{\ii k_x c_s^2
      q}{Dr} - \frac{k_x^2c_s^2}{\left(k_x^2c_s^2 + \chi
        D\right)}\frac{qc_s^2F}{Dr^2}\right]\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left\{\sigma^2\left(\frac{k_x^2}{D} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{Dr}\right)\frac{d^2\ln\rho}{dz^2} -
    \frac{c_s^2}{D}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left[\left(1-\chi\right) +
     \frac{\chi}{\left(k_x^2c_s^2 + \chi D\right)}\frac{qc_s^2 F}{r^2}\right] 
   \right\}\delta v_z,
\end{align}
where we recall $\chi = \left(1-\ii\sigma t_c\right)/\left(1-\ii\sigma
t_c \gamma\right)$. Making the low-frequency approximation with the
replacement $D\to \Omega_k^2$ gives, in terms of dimensionless
variables,
\begin{align}
   &\delta v_z ^{\prime\prime} + \left[1 + \ii\epsilon q\hat{k} -
    \frac{ \hat{k}^2}{
      \left(\hat{k}^2+\chi\right)}q \epsilon^2F\right]\ln\rho^{\prime}\delta v_z^\prime +
  \left\{\left(\chi + \ii \epsilon q
      \hat{k}\right)\ln\rho^{\prime\prime} - \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii\epsilon
      q\hat{k}\right)\left[1 - \chi +
      \frac{\chi}{\left(\hat{k}^2+\chi\right)}q\epsilon^2F\right]\right\}\delta v_z \notag\\&=
  -\hat{\sigma}^2\left(\hat{k}^2+\chi\right)\delta v_z\label{adia_diso3}.
\end{align}   
Eq. \ref{adia_diso3} differs from Eq. \ref{vertiso_gov_nondim} by terms
proportional to $\epsilon^2$. For a thin disk, $\epsilon\ll1$, so
neglecting these terms has no qualitative effect on the discussion in 
\S\ref{analytic_relax}.

@@

\subsection{Limiting form with real frequencies and $\khat^2\gg 1$}\label{disp_neut_limit}
Assuming neutral modes with real $\hat{\sigma}=\hat{\omega}$ and $\khat^2\gg 1$, the real and imaginary parts of
Eq. \ref{relax_disp} are approximately 
\begin{align}
  0 =&M(M+1)(1-\epsilon^2 q^2 \khat^2) + 4\beta\epsilon q M(M+1)\left(\hat{\omega}\khat\right) 
 +\left[1 +
    \gamma\beta^2\left(1-\gamma\right)(1+2M)^2\right]\left(\hat{\omega}\khat\right)^2
 + 2\epsilon q \gamma\beta \left(\hat{\omega}\khat\right)^3 -  \left(\hat{\omega}\khat\right)^4\notag \\
  &+ \beta^2\gamma^2\hat{\omega}^2
  \left(\hat{\omega}\khat\right)^4,\label{relax_cond1} \\
   0=& 2\epsilon q M (M+1)\khat +
   \beta(1-\gamma)(1+2M)^2\khat\left(\hat{\omega}\khat\right) 
   + \epsilon q \khat \left(\hat{\omega}\khat\right)^2 -
   2\beta\hat{\omega}\left(\hat{\omega}\khat\right)^2
   - \epsilon q
   \gamma^2\beta^2\hat{\omega}\left(\hat{\omega}\khat\right)^3 
   +
   2\beta\gamma\hat{\omega}\left(\hat{\omega}\khat\right)^4 \label{relax_cond2}. 
\end{align}
These equations are used to derive a critical thermal timescale in
\S\ref{iso_vsi_beta_crit}. For fixed $M=O(1)$,  we assume
$|\hat{\omega}\khat|$ remain constant but consider $\khat\gg 1$ so
that $|\hat{\omega}|\ll 1$.  Then for a thin disk $\epsilon \ll 1$
we neglect the second, fourth and last term in Eq. \ref{relax_cond1};
and neglect the last three terms in Eq. \ref{relax_cond2}. This
procedure gives Eq. \ref{relax_cond_simp1}---\ref{relax_cond_simp2},
but is not valid if $\beta$ or $|q|$ is large.

@@

@@

\subsubsection{Maximum growth rate}\label{max_growth1}
Here we show that the growth rate is limited by the maximum vertical
shear rate in the domain. % Writing $\hat{\sigma} = \hat{\omega} +
% \ii\hat{\nu}$ with real $\hat{\omega}$ and $\hat{\nu}$, 
The real and imaginary parts of 
Eq. \ref{integral_relation1} are
\begin{align}
  \left(\hat{\omega}^2-\hat{\nu}^2\right)\left(1+\hat{k}^2\right)
  \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z} -
  \int_{\zhat_1}^{\zhat_2}\rho|W^\prime|^2d\hat{z}
  \notag
  =\real\left[\ii
    \epsilon\hat{k}q \int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\zhat\right], \\
   2\hat{\omega}\hat{\nu}\left(1+\hat{k}^2\right)
  \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z}
  =\imag\left[\ii
    \epsilon\hat{k}q \int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right],
\end{align}
where we recall $\hat{\omega} = \real \hat{\sigma}$ and
$\hat{\nu}=\imag\hat{\sigma}$. 
Adding the square of these equations give
\begin{align}
  \left[|\hat{\sigma}|^2\left(1+\hat{k}^2\right)
    \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z} -
    \int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}\right]^2
  +4\hat{\nu}^2\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}
  =\left|\ii
    \epsilon\hat{k}q\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|^2.
\end{align}
It is clear that
\begin{align}\label{sigma_finite_domain} 
  4\hat{\nu}^2\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime
  \right|^2d\hat{z} 
  \leq\left|
    \epsilon\hat{k}q\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|^2.
\end{align}
On the left hand side of this inequality, we apply the Cauchy-Schwarz
inequality to obtain
\begin{align}
  \left( \int_{\zhat_1}^{\zhat_2}\rho
    |W|\left|W^\prime \right|d\hat{z}\right)^2\leq
  \int_{\zhat_1}^{\zhat_2}\rho 
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}.
\end{align}
On the right hand side of Eq. \ref{sigma_finite_domain} we have
\begin{align}
  \left|\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|\leq \int_{\zhat_1}^{\zhat_2}\rho
  \left|f(\hat{z})W^*W^\prime \right|d\hat{z}
  \leq
  \mathrm{max}\left(|f|\right)\int_{\zhat_1}^{\zhat_2}\rho
  |W|\left|W^\prime \right|d\hat{z},
\end{align}
where $\mathrm{max}(|f|)$ is the maximum value of $|f|$ in
$\zhat\in[\zhat_1,\zhat_2]$. Inserting these inequalities into
Eq. \ref{sigma_finite_domain} gives
\begin{align}\label{max_growth}
  |\hat{\nu}|\leq
  \frac{\epsilon |\hat{k} q|}{2\sqrt{1+\hat{k}^2}}\mathrm{max}(|f|). 
\end{align}
Recalling that $f$ represents vertical shear (Eq. \ref{fz_shear}), it
follows that the maximum possible growth rate of unstable modes,
satisfying the above boundary conditions, is limited by the maximum
vertical shear rate in the domain considered.

@@

We seek power series solutions to Eq. \ref{iso_ode3}, 
\begin{align}
  W(\zhat) = \sum_{l=0}^\infty a_l\zhat^l. 
\end{align}
Then the coefficients $a_l$ are given by the recurrence relation
\begin{align}
  (l+2)(l+1)a_{l+2} +
  \left[\hat{\sigma}^2\left(1+\frac{\khat^2}{1-\hat{\sigma}^2}\right)
    - l\left(1+\frac{\ii \epsilon q
        \hat{k}}{1-\hat{\sigma}^2}\right)\right] a_l = 0. 
\end{align}
We demand the series to terminate  at $l=L$, i.e. a polynomial of
order $L$, so that the vertical kinetic energy remain bounded as 
$|\zhat|\to\infty$.  Then the eigenfrequency $\hat{\sigma}$ is given via 
\begin{align}
\hat{\sigma}^4 - \left(L+1+\khat^2\right)\hat{\sigma}^2 + L\left(1 +
  \ii\epsilon q \khat\right) = 0. \label{simple_growth}
\end{align}

\subsubsection{Stability without vertical shear}\label{stable_novshear}
In the absence of vertical shear $q=0$, Eq. \ref{simple_growth} can be
written as \begin{align}
  \hat{\sigma}^2\khat^2 =
  \left(L-\hat{\sigma}^2\right)\left(1-\hat{\sigma}^2\right), \label{iso_disp_full}
\end{align}
which is just the dispersion relation for axisymmetric isothermal waves in a
vertically isothermal disk
\citep[e.g.][]{okazaki87,takeuchi98,tanaka02,zhang06,ogilvie13,barker14}. In 
this case the solutions are Hermite polynomials, $W\propto
\He_L(\zhat)$.  The eigenfrequency $\hat{\sigma} = \hat{\omega}$ is real and the disk
is stable. The low-frequency branch of Eq. \ref{iso_disp_full} are 
inertial waves \citep{balbus03}. For 
$|\hat{\omega}|\ll 1$ and $L\geq 1$ the dispersion relation is
$\hat{\omega}^2\khat^2 = L$, or $\hat{\omega}\propto \khat^{-1}$ for
fixed $L$.  This inverse relation has been qualitatively
observed in numerical simulations of \cite{stoll14}.
